---
title: "UP BMEC Guide to R - Introductory Programming"
author: "Sam Gianan"
date: "2026-08-31"
output:
  github_document:
    toc: true
    toc_depth: 3
    html_preview: true
    keep_html: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introductory Programming

Welcome to the first module!

R has a lot of capabilities when it comes to data analysis and machine learning, but it's first important to understand the basics.

## Understanding RStudio

### Panes

If you are using RStudio, you will notice that upon launching, there are three main sections:

+--------------+---------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Location     | Name                      | Purpose                                                                                                                                                                    |
+==============+===========================+============================================================================================================================================================================+
| Left         | Console                   | Where your live code is ran                                                                                                                                                |
+--------------+---------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Upper Right  | Environment/History/Git   | Keeps track of variables, dataframes, and commands currently stored in your computer's memory                                                                              |
|              |                           |                                                                                                                                                                            |
|              |                           | You can also commit changes to GitHub through the *Git* section                                                                                                            |
+--------------+---------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Bottom Right | Files/Plots/Packages/Help | Shows an outline of all files and packages in your project, as well as generated plots                                                                                     |
|              |                           |                                                                                                                                                                            |
|              |                           | The *Help* section may be used to search up documentation for a function without leaving the IDE through `?[function]` (try typing `?mean` in the console to test it out!) |
+--------------+---------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

### Creating Your First Project and Script

To follow along, create your first project and script in R!

1.  Click *File \> New Project... \> New Directory*.
2.  Select *New Project* as the project type.
3.  Type a name for your project.
4.  Click *Browse...* to choose a folder to save your project in.
5.  Click *Create Project*.

While you can follow along by writing code in the console, it would be better to create a file (i.e. an R script) to save and organize your code.

1.  Click *File \> New File \> R Script*, which will open the *Source* pane. This completes the final pane you'll see in a standard RStudio layout.
2.  Inside the *Source* pane, you can now type the obligatory introductory code snippet:

```{r}
print("Hello world!")
```

3.  Run the code with the `Run` button in the upper right corner of the *Source* pane.

Bonus things to know to organize your scripts:

- Commenting on your code is good practice for yourself and others to understand what your script does. To comment on R:

  - A single-line comment only needs a hashtag, `# like this`!

  - To comment out multiple lines, highlight the lines you want to turn into a comment then hit `Ctrl + Shift + C` (Windows/Linux) or `Cmd + Shift + C` (Mac).

  - To create sections in your R script, adding four or more consecutive hashtags (`####`), hyphens (`----`), or equals signs (`====`) can add collapsible headers to your script.

```{r}
# This is a single-line comment

# This multi-line comment...
# ...was made with Ctrl + Shift + C

# This makes a new section! ####

# So does this ----

# This too ====
```

## Basic Functions and Variables

We can now move on to introductory programming! To create your first variable, R actually uses the `<-` operator instead of the more common `=` operator. For example:

```{r}
x <- 1
y <- 2
```

Don't worry about semicolons at the end of a line! You can use them though to assign variables in one line:

```{r}
a <- 3; b <- 4
```

To delete a variable, we use the `rm()` function:

```{r}
rm(y)

# To delete more than one object, just separate them with a comma!

rm(a, b)

# To delete everything in your environment:

rm(list = ls()) 
```

Re-initialize `x` and `y` to use for the next few examples.

Operations can be done in R as demonstrated:

```{r}
x <- 1
y <- 2

x + y # Addition
y - x # Subtration
x * y # Multiplication
x / y # Division
y %% x # Modulus
```

## Data Types

Before creating data structures, it is important to first know core data types in R. There are six atomic data types, which are the fundamental datatypes we will be working with.

### Character

This data is meant for text. To create a value of the character data type, simply wrap the text (or even number) in single or double quotes.

```{r}
my_character1 <- "hello"
my_character2 <- "101"
```

To convert from one data type to the character data type, we use `as.character()`.

```{r}
my_character3 <- as.character(67)
print(my_character3)
```

Notice that the number has quotes around it!

### Numeric

This acts as the default type for any number, whether or not it has a decimal.

```{r}
my_numeric1 <- 3.14
my_numeric2 <- 150
```

We use `as.numeric()` to convert to a numeric data type; we'll use this to convert `my_character3` back into numeric data.

```{r}
my_numeric3 <- as.numeric(my_character3)
print(my_numeric3)
```

The quotes have now disappeared.

### Integer

Integers are for data without decimal places.

```{r}
my_integer1 <- 120L
my_integer2 <- 50L
```

Once again, we can use `as.integer()` to convert `my_character2`.

```{r}
my_integer3 <- as.integer(my_character2)
print(my_integer3)
```

When dividing integers, the final value will be in the numeric data type. However, you can specify to fetch the quotient in integer form with integer division.

```{r}
# Normal division (returns a numeric value)
my_integer1 / my_integer2

# Integer division (returns an integer value)
my_integer1 %/% my_integer2

# Remainder (Returns an integer value)
my_integer1 %% my_integer2
```

To get the floor and ceiling values, we use, well, `floor()` and `ceiling()`.

```{r}
floor(my_numeric1)
ceiling(my_numeric1)
```

### Logical

This is made of true or false statements (in uppercase), and will be used during data manipulation and filtering later on.

```{r}
is_weekday <- TRUE
is_weekend <- FALSE
print(is_weekday)
print(is_weekend)

# Shortcuts can also be used
is_daytime <- T
is_nighttime <- F
print(is_daytime)
print(is_nighttime)
```

More than often though, the way to generate logical values is by comparing values through logical operators. We'll go into more detail on this later, but here is a short demonstration just to show the capabilities:

```{r}
# Check if values are equal
5 == 10
5 != 10

# Check for size comparison
5 < 10
5 <= 10
5 >= 10
5 > 10

# Check if in a vector
"Red" %in% c("Red", "Orange", "Yellow")
```

### Complex

The complex data type is used to store complex numbers, but this won't really be used.

```{r}
# Storing real + imaginary value
my_complex1 <- 5 + 6i
print(my_complex1)
# Storing just imaginary value
my_complex2 <- 7i
print(my_complex2)
```

### Raw

The raw data type is meant to store binary data, and is meant for low-level processing.

```{r}
# Generating raw vector of length 10
my_raw1 <- raw(10)
print(my_raw1)

# Converting text to raw bytes and vice versa
my_raw2 <- charToRaw(my_character1)
print(my_raw2)
rawToChar(my_raw2)
```

## Data Structures

### Vectors

To create a vector, we use c(), which means to combine. A vector must have the same data type (i.e. all numeric, all logical, all dates, etc.). Vectors come in handy for data cleaning and manipulation, such as for searching and filtering.

```{r}
fruits <- c("Apple", "Orange", "Banana")
fruits
```

To add a new value, you may either use `c()` or the `append()` function. `append()` is more flexible, as it allows you to add a value in the middle of a vector, not just at the beginning or at the end.

```{r}
# Using c()
fruits <- c(fruits, "Grapes") # Adds at end of a vector
fruits
fruits <- c("Blueberry", fruits) # Adds at the beginning of a vector 
fruits

# Using append()
fruits <- append(fruits, "Kiwi") # Appends at the end of a vector
fruits
fruits <- append (fruits, "Strawberry", after = 3) # Appends after 3 elements, making it the fourth item in the vector
fruits
```

We'll go more in-depth on what vectors could be used for once we get into data cleaning and manipulation.

### Lists

Unlike vectors, lists may hold more than one data type. We can create them with the `list()` function. Lists can also hold vectors, matrices, dataframes, and even other lists.

```{r}
student <- list(
  name = "John Doe",
  age = 18,
  birthday = as.Date("2004-08-13", # We'll go more into the as.Date() function in the next module
  taken_ges = c("Eng 13", "Kas 1", "Speech 30"),
  has_graduated = FALSE
))

str(student) # For printing the list
```

As observed, various data types can be stored within the student list.

### Matrices

Matrices are two-dimensional by nature with rows and columns. To create a matrix, you can either (1) use an existing vector and divide it by column or row, or (b) combine vectors with each acting as a row or column. The latter is done with `cbind()` or `rbind()`, and can be used on vectors with different lengths.

```{r}
# Vectors for demonstration
vec_1 <- c(1, 2, 3, 4, 5, 6, 7, 8, 9)
vec_2 <- c(1, 2, 3)
vec_3 <- c(1, 2, 3, 4, 5) 

# To fill up a matrix by column
matrix_col <- matrix(vec_1, nrow = 3, ncol = 3)
print(matrix_col) 

# To fill up a matrix by row
matrix_row <- matrix(vec_1, nrow = 3, ncol = 3, byrow = TRUE)
print(matrix_row)
```

Notice the difference in order of how the matrices are set up. Furthermore, by default, the byrow argument in the matrix() function is set to FALSE, so when filling up matrices by column, you do not need to specify its value.

```{r warning = TRUE}
matrix_cbind <- cbind(vec_2, vec_3)
print(matrix_cbind) 

matrix_rbind <- rbind(vec_2, vec_3)
print(matrix_rbind)
```

You can notice that since `vec_2` is smaller than `vec_3`, the function loops the smaller vector's values to fill up the remaining spaces. Furthermore as the number of values in `vec_3` is not divisible by the number of values in `vec_2` (i.e. 5 divided by 3 has a remainder), a warning will be given, as all of the values in `vec_2` are not looped through.

If you prefer though that these values are not looped and are instead NA values, you can set the length of the smaller vector to that of the larger one, which will cause the placement of NA values.

```{r}
max_len <- max(length(vec_2), length(vec_3))
length(vec_2) <- max_len
cbind(vec_2, vec_3)
```

### Dataframes

Dataframes are the structure that most closely resemble spreadsheets; they are two-dimensional and can hold more than one data type.

```{r}
name_vec <- c("Elijah", "Hannah", "John")
age_vec <- c(18, 20, 19)
university_vec <- c("University of the Philippines - Diliman", "De La Salle University", "University of Santo Tomas")

my_df <- data.frame(
  name_vec, age_vec, university_vec
)

print(my_df)
```

This is the type you'll be working with when handling actual data from existing datasets!

### Class Identification

To check for the classification of values, we can either use `class()` or `typeof()`. Using `class()` gives us the high-level, general data type, while `typeof()` gives the low-level, internal data type. This could be better understood through a demonstration.

```{r}
type_test <- data.frame(id = 1:2, name = c("A", "B"))
class(type_test)
typeof(type_test)
```

We can see that running `class()` gives us "data.frame", letting us know that R treats type_test as a dataframe. However, a dataframe is also just a list of vectors, which is the value given by `typeof()`.

------------------------------------------------------------------------

# This module is still under construction!

Changes were pushed to test committing in GItHub.

Don't worry, the rest of the module will be up soon!

------------------------------------------------------------------------

## License and Copyright

Copyright (c) 2026 UP Biomedical Engineering Circle.

This project is licensed under the terms of the MIT license.
